\mythesistitleNOTSYNPOSIS \tnote[fyi]{Synopsis title is ``Generalization of Consecutive Ones Property of Binary Matrices''}

### 2012-02-02 src/020chap2.tex
-- sent to n for review
- moved ded to quot. added ded.
- pqtree, pctree figures completed.
- fixed a few missing bibs
- random clean up including temptexts. see diff.

This chapter surveys several results that are significant to this
thesis or to \COP in general. These predominantly pertain to
characterizations of \COP, algorithmic tests to check for \COP (COT),
optimization problems on binary matrices that do not have \COP and
some applications of \COP.

\tnote[dire]{ADD: have a few lines about organization of chapter}

\section{\COP in Graph Theory}

@@

\begin{definition}{\emph{Graphs that relate to
      COP.\cite[Def.~2.5]{d08phd}}} %
  \label{def:graphwithcop} %
  Let $G$ be a graph and $H$ be a bipartite graph.
  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \emph{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \emph{concave-round} if its
    augmented adjacency matrix has \CROP. \tnote[pressing]{cite BHY00}
  \item $G$ is an \emph{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \tnote[pressing]{cite Ben59,
      Haj57}.  $G$ is an interval graph \iff its maximal clique matrix
    has COP \cite{fg65}\footnote{This follows \cite{gh64} which states
      that the maximal cliques of interval graph $G$ can be linearly
      ordered \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.}
    \begin{enumerate}[a.]
    \item $G$ is a \emph{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.\footnotemark[2]
    \item $G$ is a \emph{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.\footnote{The
        set of unit interval graphs and the set of proper interval
        graphs are the same} \tnote[pressing]{cite {rob69,gar07} in
        footnote. pg 33 dom}
    \end{enumerate}
 \item $G$ is a \emph{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.\tnote[pressing]{how is
      CO/ROP related?}
  \item $H$ is \emph{convex bipartite on columns (rows)} if its half
    adjacency matrix has \COP on rows (columns).%
    \label{def::convexbi}
  \item $H$ is \emph{biconvex bipartite} or \emph{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \emph{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

@@

The first polynomial time algorithm for \COP testing was by
\cite{fg65} which uses overlapping properties of columns with \un
s. Their result has close relations to the characterization of
interval graphs by \cite{gh64}. A graph $G$ is an interval graph \iff
all its maximal cliques can be linearly ordered \stt for any vertex
$v$ in $G$, all the cliques that $v$ is incident on are consecutive in
this order. Clearly, this means that the maximal clique incidence
matrix\footnote{
  Definition~\ref{def:graphmatrices}~\ref{def::maxcliquematrix}} must
have \COP on rows.

@@

Booth and Lueker in their paper \cite{bl76} gave the first linear
algorithm\footnote{Time complexity is $O\left(m+n+f\right)$ where $m
  \times n$ is the order of the input matrix and $f$ is the number of
  $\un$s in it.} for \cop testing while given a linear time interval
graph recognition algorithm by a simplification of \tnote[important]{cite -
  A. Lempel, S. Even and I. Cederbaum, An Algorithm for Planarity
  Testing of Graphs, Theory of Graphs, ed., P. Rosenstiehl, Gordon and
  Breach, New York, (1967), 215-232.  }'s planarity test algorithm.
% Hsu in wlh01 said BL76 gave a planarity test - "the induced PQ-tree
% algorithm can considerably simplify Booth and Lueker’s modification
% of Lempel, Even and Cederbaum’s planarity test. " 
%
% McConnell in hm03 said the first linear algo for interval graph
% recog was by BL76 - "Benzer’s work and other applications of
% interval graphs motivated a search for eﬃcient algorithms for
% recognizing interval graphs, and for constructing a set of intervals
% to represent the graph when it is one [8]. A linear-time algorithm
% was given in 1976 by Booth and Lueker [2]." 
\cite{bl76} introduces a data structure called \PQtree and their \COP
testing algorithm is a constructive one that outputs a \PQtree if the
input has \COP. A \PQtree represents all the \COP orderings of the
matrix it is associated with. \cite{bl76}'s algorithm uses the fact
that if a matrix has \COP, a \PQtree for it can be constructed. It is
interesting to note that aside from interval graph recognition and
\COP testing, \PQtree is also useful in other applications like
finding planar embeddings of planar graphs \cite{lec67,mcc04} and
recognizing \CROP in a matrix.

%\begin{minipage}[h]{5.5in}
\begin{definition}[\PQtree \cite{bl76, mcc04}]
  A \PQtree of matrix $M$ with \COP on columns (rows), is a tree with
  the following properties.
  \begin{enumerate}[i.]
    \singlespacing
  \item Each leaf uniquely represents a row (column) of $M$. The leaf
    order of the tree gives a \COP order for column (row)\footnote{Note
      that \COP order for column requires permutation of rows and vice
      versa.} for $M$.
  \item Every non-leaf node in the tree is labeled $P$ or $Q$.
  \item \label{def::nodep} The children of $P$ nodes are
    unordered. They can be permuted in any fashion to obtain a new
    \COP order for $M$.
  \item \label{def::nodeq} The children of $Q$ nodes are linearly
    ordered. Their order can be reversed to obtain a new \COP
    order for $M$.
  \end{enumerate}
  \label{def:pqtree}
\end{definition}
%\end{minipage}

@@

After the invention of \PQtrees, presumably due to the implementation
challenge it posed, there has been several variants of the same in the
literature, like \PCtree \cite{sh99,wlh01,hm03}, generalized \PQtree
\cite{km89,mcc04}, \PQRtree \cite{mm96,mpt98} etc. Most of these are
generalizations of \PQtree\ -- for instance, \PCtree is generalized to
matrices with \CROP, \PQRtree and generalized \PQtree are generalized
to matrices and set systems with or without \COP. \cite{km89} invented
a modified form of \PQtree a simpler incremental update of the tree
only for recognizing interval graphs.  \cite{kr88} constructed
efficient parallel algorithms for manipulating \PQtrees. \temptext{dom
  chapter 2 pg 40 Variations of PQTrees first para - summarize.}

In the next few sections we will see some of these
variations.
%Their algorithm was improved by \cite{tm05}
%\tnote{improved in terms of what?}. 
% describes an improved algorithm to
% build PQR trees.

@@

Section~\ref{sec:motive} mentions how a binary matrix naturally maps
to a system of sets.  A set can be constructed for each column of
matrix with its elements being those row indices at which the column
has \un s. Thus the collection of sets corresponding each column of
the matrix forms a set system with universe as the set of all row
indices of the matrix.  This simple construction is formally described
in Definition~\ref{def:matrixsetsystem} along with the idea of \cop
for set systems\footnote{As seen in Section~\ref{sec:courseschedule}}.

\begin{definition}[{\em \Cop for set systems}]%
  \label{def:matrixsetsystem}%
  % [\emph{Set system of a binary matrix}.]
  Let $M$ be a binary matrix of order $n \times m$ and $\{c_i \mid i
  \in [m]\}$ be the columns in $M$.  A set system $\cF_{M} = \{S_i
  \mid S_i \subseteq [n], i \in [m]\}$ is defined such that for every
  column $c_i$ of $M$, set $S_i = \{j \mid m_{ji} = \un \}$. The
  collection $\cF_M$ is the {\em set system of binary matrix} $M$. The
  {\em binary matrix for set system} $\cF$ is conversely constructed
  and denoted by $M^\cF$. Thus, $M^{\cF_M} = M$.%
  \par\noindent%
  A set system $\cF$ from universe $U$, $|U| = n$ has the {\em \cop}
  if there exists a linear order or permutation $\sigma = w_1w_2\ldots
  w_n$ that can be applied to $U$ \stt each set $S \in \cF$ becomes a
  consecutive subsequence\footnote{Also termed an {\em interval}}
  $w_{i}w_{i+1}\ldots w_{i+k-1}$ on $\sigma$ for some positive integer
  $i \le n+1-k$ where $k = |S|$.
  \par\noindent%
  The set $\valid{U,\cF}$ respresents all \COP orders of $\cF$ in $U$.
  \dstop
\end{definition}

@@

\begin{definition}[\em Orthogonal sets \cite{n89, mm96,
    mcc04}\footnote{\cite{mcc04} does not use the term ``mutually
    orthogonal'' but refers to the same idea as ``sets that do not
    overlap''. This terminology is also used in other literature like
    \cite{nsnrs09,wlh02}.} etc.]
  \label{def:orthosets}%
  Let $\cF$ be a set system with universe $U$ and sets $A, B \in \cF$.
  \begin{enumerate}
  \item $A$ and $B$ are said to have a {\em trivial intersection} if
    $A \cap B$ is one of the following.
    \begin{enumerate}[i.]
    \singlespacing
    \item $\emptyset$
    \item $A$
    \item $B$\footnote{In other words, $A$ and $B$ are either disjoint
        or one is the subset of the other.}
    \end{enumerate}
    % \vspace{-5mm}%

@@

\temptext{narrative is abrupt from here. add some explanation for the
  theorem and proposition below?}

@@

Thus we have a recursive algorithm that can compute the \PQRtree for
$\cF$ provided we find an element from $\ff$ in each iteration. The
non-empty sets in $\ff$ are called {\em node sets} since they form the
nodes in the \PQRtree, as alluded to by
Corollary~\ref{cor:pqrrecursive}. They are calculated as follows.
First method is by computing the overlap components of $\cF$. The
overlap components is the partition that results from the {\em
  overlap} equivalence relation which is nothing but {\em
  non-orthogonal} equivalence relation\footnote{It is easy to verify
  that this relation is indeed an equivalence relation.}. Overlap
components are linearly computable\cite{mm95,wlh92}. Once these
elements are factored out, the rest of the node sets are obtained by
the second method which is identifying {\em twin}
elements\footnote{\cite[Sec.~3]{kklv10} calls this {indistinguishable}
  elements. The equivalence class is called a {\em slot}.}. Two
elements $a, b \in U$ are twins if their membership in every set of
$\cF$ is in tandem with each other, \ie $\{a,b\} \bot \cF$. This is
clearly an equivalence relation and their equivalence classes is known
to be computable in linear time\cite{wlh01,mm96-ref11} and even in
logspace\cite{kklv10}.

@@

Theorem~\ref{th:pqrfinalmm} and the theory leading to it is very
similar to \cite[esp. Th.~2.1, 3.5. also Th.~3.2, 3.3, 3.4]{mcc04}
which categorizes the nodes in the above the three cases as {\em
  prime}, {\em linear} and {\em degenerate}
respectively. \cite{mcc04}'s generalized \PQtree is created in similar
ways as \PQRtree above. This tree is, in essence, the Hasse diagram of
what they call {\em strong elements} of weak closure,
$\cW(\cF)$\footnote{\ie $\overline{\cF}$}. Strong elements of a set
family are elements that do not overlap with any other elements in the
family, \ie it is orthogonal to all other sets\cite[Def.~3.3]{mcc04}.

@@

    &
%     \parbox[t]{\colwidth}
%       {}
    \temptext{PQR tree def. }
    &
    \parbox[t]{\colwidth}
    {
      Decomposition tree of $\cW(\cF)$, $T(\cW(\cF))$
      \temptext{PQR tree def. }
    }
    \\\hline

@@

    &
    \parbox[t]{\colwidth}
    {
      $P$-node $\Leftrightarrow \cF = \cT(\cF)$\\
      $Q$-node $\Leftrightarrow \cF = consec(\alpha)$\\
      $R$-node $\Leftrightarrow \cF = \power{U}$\\
    }&
    \temptext{give def of P Q R nodes}
    \\\lasthline

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.4]{../img/pctree_hm03.png}

  \centering
  \includegraphics[scale=0.4]{../img/pctree_dom.png}
  \caption{\figtabsize \PCtree \temptext{ PLACEHOLDER 
IMGS \cite{hm03, d08phd}}}
  \label{fig:pctree}
\end{figure}

@@

\subsection{ICPIA - a set cardinality based \COP test}
\label{sec:icpiasurvey}
\temptext{narrative is abrupt}

@@

The necessity condition of Theorem~\ref{th:icpia1} is fairly
obvious. It turns out that it is sufficient too.  This is demostrated
by two algorithms which work as follows. The first one iterates over
the assignment $\Pi$ %set system $\cF$
and changes it in each iteration till the altered
assignment %set system
has no more overlapping intervals\footnote{Intervals are the second
  element in each ordered pair of $\Pi$. Def.~\ref{def:icpia}} %sets
\cite[Alg.~1]{nsnrs09}. This is acheived by replacing two overlapping
assignment pairs $(A_i, B_i)$ and $(A_j, B_j)$ with the partitions of
$(A_i \cup A_j,B_i \cup B_j)$ induced by their overlaps -- $(A_i
\setminus A_j, B_i \setminus B_j)$, $(A_i \cap A_j, B_i \cap B_j)$ and
$(A_j \setminus A_i, B_j \setminus B_i)$.

To see it in terms of the theory explained in
Section~\ref{sec:surveycertalgo}, this results in the weak closure of
$\cF$. Let $\cF$, $\cI$ be the set system and interval
system\footnote{A set system where all the sets are intervals.},
respectively, involved in the assignment $\Pi$ and $\cF'$, $\cI'$ be
the same for the output of the algorithm $\Pi'$.  It can be observed
that $\cF' \cup \cF$ is $\overline{\cF}$ or the weak closure of
$\cF$. The analog is true for $\cI \cup \cI'$.  Moreover, this output
$\cF'$, $\cI'$ contains the strong elements of $\overline{\cF}$,
$\overline{\cI}$ respectively. It turns out that $\cI'$ is also an
interval system and that $\Pi'$ is an ICPIA \cite[Lem.~2]{nsnrs09}.

@@

What we saw so far in this section is checking for the feasibility of
a given interval assignment. ICPIA can also be used to find an
interval assignment when only the set system $\cF$ is given. Part of
the approach is similar to the overlap component idea in \cite{fg65,
  wlh02}. Once the overlap components are factored out, they can be
independently assigned a subinterval to which the sets in the
component need to be mapped. This is simple because the overlap
components are disjoint from each other. Now the subproblem is finding
\COP order for each overlap component. In each component, the set
whose overlaps with the rest of the sets in that component forms a
single inclusion chain (see Definition~\ref{def:poset}. In this
particular reference, the set of sets is the set of overlaps.)  is
chosen first and assigned the leftmost\footnote{By
  convention. Rightmost can also be chosen analogously to get a
  different \COP order.} interval. The next candidate set is chosen
from this set's overlapping sets. The interval assigned is calculated
by simple intersection cardinality means\footnote{Assume that the next
  set overlaps on one of the ``ends'' of the current set and calculate
  the interval by shifting left or right of the current interval and
  adjusting its size to match the next set's cardinality.}. Once the
next set is assigned an interval, the current assigned sets are all
checked for ICPIA. This is a backtracking algorithm -- if at any point
ICPIA fails, another overlapping set is chosen. If all overlapping set
attempts fail ICPIA, a different ``first set'' is chosen. This is a
remarkably simple algorithm due to its obvious simplicity in
implementation.

@@

\temptext{ abrupt: Add a section closing para. does this method (above)
  lead to efficient algo? }

\section{Optimization problems in COP}
\label{sec:surveycopopt}

\COP is a very beneficial property since it simplifies the structure
of the input\footnote{Binary matrix or set system.} leading to
efficient algorithms to otherwise hard problems.  While
Section~\ref{sec:surveycoptest} discusses matrices with \COP and how
to compute \COP orders, this section describes the problems that are
of interest when a matrix does not have this property.  The central
questions in this area are (i) how close is the matrix to having \COP,
mainly in terms of Tucker's forbidden submatrices, and (ii) how
optimally can one alter the matrix to attain \COP.  With regard to the
latter question, recent literature indicate that there has been a lot
of interest in matrix modification problems to make a matrix have \COP
\cite{hg02, tz07}. However, to tackle this problem it is more than
insightful to solve question (i).

To study the smallest hamming distance \temptext{(?)} between the
input matrix and one with \COP, identifying the Tucker's forbidden
substructures\footnote{See Section~\ref{sec:tucker} for detailed
  discussion on Tucker's forbidden substructures for matrices with
  \COP.} is of prime importance.

\subsection{Finding forbidden submatrices}
It is known from Theorem~\ref{th:tuckercop}
that when a matrix does not have \COP, it must have one of Tucker's
forbidden submatrices. However, none of the \COP test algorithms discussed in
Section~\ref{sec:surveycoptest} outputs the forbidden submatrix when
the test is negative. This certainly gives a hint to how hard it is to find
these substructures in the input. \temptext{Fill neg results on
  finding forbidden submartrices -- boo75, some highlighted stuff from v85}

@@

\begin{corollary}[{\cite[Cor.~3.1]{d08phd}}]
  \label{cor:tuckertheorem}
  A matrix $M$ has \COP \iff its representing bipartite graph
  $G_M$\footnote{Graph whose half adjacency matrix
    (Def.~\ref{def:xxxx}) is $M$.} does not contain an asteroidal
  triple whose three vertices correspond to columns of M.
\end{corollary}

@@

\temptext{how does it give clues
  about the $R$ node? See dom.}

@@

\temptext{\textbf{REST OF THE SUBSECTIONS}}

\temptext{ -- sect 4.1 in cite:d08phd has many results surveyed. hardness
  results, approx. results. results are usually for a class of
  matrices $(a,b)$ where number {\un}s in columns and rows are
  restriced to $a$ and $b$ . -- problem of flipping at most $k$
  entries of $M$ to make it attain COP. this is NP complete
  cite:b75-phd%
\\(1) scite:lb62 showed that interval graphs are chordal and
  AT-free. %
\\couple of lines refering to tucker's submatrices. refer
  earlier section.

@@

\section{COP in Graph Isomorphism}
\label{sec:surveygraphiso}
\tnote{(canonization theme)}

\tnote[pressing]{ADD: peo exists iff chordal. lexicographic BFS
  [tag:chordalGraph]} %

\temptext{
The survey from kklv10 conclusion.
}

>>>>

This chapter surveys several results that are significant to this
thesis or to \COP in general. These predominantly pertain to
characterizations of \COP, algorithmic tests to check for \COP,
optimization problems on binary matrices that do not have \COP and
some applications of \COP.

The chapter is organized as follows. Section~\ref{sec:copgraphtheory}
discusses the close connections of matrices with \COP in graph
theory. Section~\ref{sec:surveycoptest} discusses the important
results in the area of \cop testing (COT) -- we describe the history
of this problem's several solutions including Tucker's forbidden
submatrices, \PQtree, \PQRtree and ICPIA. \tnote[dire]{mention the section
  that compares pqr and gen pq tree} 
Section~\ref{sec:surveycopopt} briefly touches upon the scenario of
algorithmically verifying that a matrix indeed does not have \COP. Its
significance lies in the fact that almost all \COT algorithms, upon
detection of no \COP, simply reports the same without any certificate
for verification during algorithm implementation. The two approaches
presented in this section gives a certificate.

\section{\COP in Graph Theory}
\label{sec:copgraphtheory}

@@

\begin{definition}{\emph{Graphs that relate to
      COP.\cite[Def.~2.5]{d08phd}}} %
  \label{def:graphwithcop} %
  Let $G$ be a graph and $H$ be a bipartite graph.
  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \emph{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \emph{concave-round} if its
    augmented adjacency matrix has \CROP. \tnote[pressing]{cite BHY00}
  \item $G$ is an \emph{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \tnote[pressing]{cite Ben59,
      Haj57}.  $G$ is an interval graph \iff its maximal clique matrix
    has COP \cite{fg65} (This follows \cite{gh64} which states
      that the maximal cliques of interval graph $G$ can be linearly
      ordered \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.)
    \begin{enumerate}[a.]
    \item $G$ is a \emph{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.
    \item $G$ is a \emph{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.
    \end{enumerate}
    The set of unit interval graphs and the set of proper interval
    graphs coincide \cite{rob69, gar07}.
 \item $G$ is a \emph{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.\tnote[pressing]{how is
      CO/ROP related?}
  \item $H$ is \emph{convex bipartite on columns (rows)} if its half
    adjacency matrix has \COP on rows (columns).%
    \label{def::convexbi}
  \item $H$ is \emph{biconvex bipartite} or \emph{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \emph{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

@@

The first polynomial time algorithm for \COP testing was by
\cite{fg65} which uses overlapping properties of columns with \un
s. Their result has close relations to the characterization of
interval graphs by \cite{gh64}. A graph $G$ is an interval graph \iff
all its maximal cliques can be linearly ordered \stt for any vertex
$v$ in $G$, all the cliques that $v$ is incident on are consecutive in
this order. Clearly, this means that the maximal clique incidence
matrix (see
Definition~\ref{def:graphmatrices}~\ref{def::maxcliquematrix}) must
have \COP on rows.

@@

Booth and Lueker in their paper \cite{bl76} gave the first linear
algorithm for \cop testing while given a linear time interval graph
recognition algorithm by a simplification of \cite{lec67}'s planarity
test algorithm. This \COT algorithm has time complexity
$O\left(m+n+f\right)$ where $m \times n$ is the order of the input
matrix and $f$ is the number of $\un$s in it.
% Hsu in wlh01 said BL76 gave a planarity test - "the induced PQ-tree
% algorithm can considerably simplify Booth and Lueker’s modification
% of Lempel, Even and Cederbaum’s planarity test. " 
%
% McConnell in hm03 said the first linear algo for interval graph
% recog was by BL76 - "Benzer’s work and other applications of
% interval graphs motivated a search for eﬃcient algorithms for
% recognizing interval graphs, and for constructing a set of intervals
% to represent the graph when it is one [8]. A linear-time algorithm
% was given in 1976 by Booth and Lueker [2]." 
\cite{bl76} introduces a data structure called \PQtree and their \COP
testing algorithm is a constructive one that outputs a \PQtree if the
input has \COP. A \PQtree represents all the \COP orderings of the
matrix it is associated with. \cite{bl76}'s algorithm uses the fact
that if a matrix has \COP, a \PQtree for it can be constructed. It is
interesting to note that aside from interval graph recognition and
\COP testing, \PQtree is also useful in other applications like
finding planar embeddings of planar graphs \cite{lec67,mcc04} and
recognizing \CROP in a matrix.

%\begin{minipage}[h]{5.5in}
\begin{definition}[\PQtree \cite{bl76, mcc04}]
  A \PQtree of matrix $M$ with \COP on columns (rows), is a tree with
  the following properties.
  \begin{enumerate}[i.]
    \singlespacing
  \item Each leaf uniquely represents a row (column) of $M$. The leaf
    order of the tree gives a \COP order for column (row) (\COP order
    for column requires permutation of rows and vice versa.) for $M$.
  \item Every non-leaf node in the tree is labeled $P$ or $Q$.
  \item \label{def::nodep} The children of $P$ nodes are
    unordered. They can be permuted in any fashion to obtain a new
    \COP order for $M$.
  \item \label{def::nodeq} The children of $Q$ nodes are linearly
    ordered. Their order can be reversed to obtain a new \COP
    order for $M$.
  \end{enumerate}
  \label{def:pqtree}
\end{definition}
%\end{minipage}

@@

After the invention of \PQtrees, presumably due to the implementation
challenge it posed, there has been several variants of the same in the
literature, like \PCtree \cite{sh99,wlh01,hm03}, generalized \PQtree
\cite{km89,mcc04}, \PQRtree \cite{mm96,mpt98} etc. Most of these are
generalizations of \PQtree\ -- for instance, \PCtree is generalized to
matrices with \CROP, \PQRtree and generalized \PQtree are generalized
to matrices and set systems with or without \COP. \cite{km89} invented
a modified form of \PQtree, called $MPQ$-trees, a simpler incremental
update of the tree only for recognizing interval graphs.  \cite{kr88}
constructed efficient parallel algorithms for manipulating \PQtrees.

In the next few sections we will see some of these variations.

@@

Section~\ref{sec:motive} mentions how a binary matrix naturally maps
to a system of sets.  A set can be constructed for each column of
matrix with its elements being those row indices at which the column
has \un s. Thus the collection of sets corresponding each column of
the matrix forms a set system with universe as the set of all row
indices of the matrix.  This simple construction is formally described
in Definition~\ref{def:matrixsetsystem} along with the idea of \cop
for set systems (As seen in Section~\ref{sec:courseschedule}).

\begin{definition}[{\em \Cop for set systems}]%
  \label{def:matrixsetsystem}%
  % [\emph{Set system of a binary matrix}.]
  Let $M$ be a binary matrix of order $n \times m$ and $\{c_i \mid i
  \in [m]\}$ be the columns in $M$.  A set system $\cF_{M} = \{S_i
  \mid S_i \subseteq [n], i \in [m]\}$ is defined such that for every
  column $c_i$ of $M$, set $S_i = \{j \mid m_{ji} = \un \}$. The
  collection $\cF_M$ is the {\em set system of binary matrix} $M$. The
  {\em binary matrix for set system} $\cF$ is conversely constructed
  and denoted by $M^\cF$. Thus, $M^{\cF_M} = M$.%
  \par\noindent%
  A set system $\cF$ from universe $U$, $|U| = n$ has the {\em \cop}
  if there exists a linear order or permutation $\sigma = w_1w_2\ldots
  w_n$ that can be applied to $U$ \stt each set $S \in \cF$ becomes a
  consecutive subsequence (slso termed an {\em interval})
  $w_{i}w_{i+1}\ldots w_{i+k-1}$ on $\sigma$ for some positive integer
  $i \le n+1-k$ where $k = |S|$.
  \par\noindent%
  The set $\valid{U,\cF}$ respresents all \COP orders of $\cF$ in $U$.
  \dstop
\end{definition}

@@

\begin{definition}[\em Orthogonal sets \cite{n89, mm96,
    mcc04}\footnote{\cite{mcc04} does not use the term ``mutually
    orthogonal'' but refers to the same idea as ``sets that do not
    overlap''. This terminology is also used in other literature like
    \cite{nsnrs09,wlh02}.} etc.]
  \label{def:orthosets}%
  Let $\cF$ be a set system with universe $U$ and sets $A, B \in \cF$.
  \begin{enumerate}
  \item $A$ and $B$ are said to have a {\em trivial intersection} if
    $A \cap B$ is $\emptyset$ or $A$ or $B$.  In other words, $A$ and
    $B$ are either disjoint or one is the subset of the other.

@@

\tnote[dire]{narrative is abrupt from here. add some explanation for the
  theorem and proposition below?}

@@

Thus we have a recursive algorithm that can compute the \PQRtree for
$\cF$ provided we find an element from $\ff$ in each iteration. The
non-empty sets in $\ff$ are called {\em node sets} since they form the
nodes in the \PQRtree, as alluded to by
Corollary~\ref{cor:pqrrecursive}. They are calculated as follows.
First method is by computing the overlap components of $\cF$. The
overlap components is the partition that results from the {\em
  overlap} equivalence relation which is nothing but {\em
  non-orthogonal} equivalence relation (It is easy to verify
  that this relation is indeed an equivalence relation.). Overlap
components are linearly computable\cite{mm95,wlh92}. Once these
elements are factored out, the rest of the node sets are obtained by
the second method which is identifying {\em twin}
elements\footnote{\cite[Sec.~3]{kklv10} calls this {indistinguishable}
  elements. The equivalence class is called a {\em slot}.}. Two
elements $a, b \in U$ are twins if their membership in every set of
$\cF$ is in tandem with each other, \ie $\{a,b\} \bot \cF$. This is
clearly an equivalence relation and their equivalence classes is known
to be computable in linear time\cite{wlh01,mm96-ref11} and even in
logspace\cite{kklv10}.

@@

Theorem~\ref{th:pqrfinalmm} and the theory leading to it is very
similar to \cite[esp. Th.~2.1, 3.5. also Th.~3.2, 3.3, 3.4]{mcc04}
which categorizes the nodes in the above the three cases as {\em
  prime}, {\em linear} and {\em degenerate}
respectively. \cite{mcc04}'s generalized \PQtree is created in similar
ways as \PQRtree above. This tree is, in essence, the Hasse diagram of
what they call {\em strong elements} of weak closure,
$\cW(\cF)$(\ie $\overline{\cF}$). Strong elements of a set
family are elements that do not overlap with any other elements in the
family, \ie it is orthogonal to all other sets\cite[Def.~3.3]{mcc04}.

@@

    &
%     \parbox[t]{\colwidth}
%       {}
    \tnote[dire]{PQR tree def. }
    &
    \parbox[t]{\colwidth}
    {
      Decomposition tree of $\cW(\cF)$, $T(\cW(\cF))$
      \tnote[dire]{PQR tree def. }
    }
    \\\hline

@@

    &
    \parbox[t]{\colwidth}
    {
      $P$-node $\Leftrightarrow \cF = \cT(\cF)$\\
      $Q$-node $\Leftrightarrow \cF = consec(\alpha)$\\
      $R$-node $\Leftrightarrow \cF = \power{U}$\\
    }&
    \tnote[dire]{give def of P Q R nodes}
    \\\lasthline

@@

\begin{figure}[htbp]
  \centering
  %\includegraphics[scale=0.4]{../img/pctree_hm03.png}

  \begin{tabular}[h]{lcc}
    \pctreegraphmatrix & 
    \pctreegraphi &
    \pctreegraphii
  \end{tabular}
  \caption[\figtabsize \PCtree]{\figtabsize \PCtree with the
    associated matrix. The dark circular nodes are $P$-nodes and the
    double circular nodes are $C$-nodes.  The two \PCtrees are both
    for the same matrix. The second tree is obtained by performing the
    ``mirroring'' operation on the subtree (highlighted in the figure)
    rooted at the bottom right $C$-node.  \cite[Fig.~2.11]{d08phd}}
  \label{fig:pctree}
\end{figure}

@@

\subsection{ICPIA - a set cardinality based \COP test}
\label{sec:icpiasurvey}
\tnote[dire]{narrative is abrupt}

@@

The necessity condition of Theorem~\ref{th:icpia1} is fairly
obvious. It turns out that it is sufficient too.  This is demostrated
by two algorithms which work as follows. The first one iterates over
the assignment $\Pi$ %set system $\cF$
and changes it in each iteration till the altered
assignment %set system
has no more overlapping intervals (intervals are the second
  element in each ordered pair of $\Pi$. Def.~\ref{def:icpia}) %sets
\cite[Alg.~1]{nsnrs09}. This is acheived by replacing two overlapping
assignment pairs $(A_i, B_i)$ and $(A_j, B_j)$ with the partitions of
$(A_i \cup A_j,B_i \cup B_j)$ induced by their overlaps -- $(A_i
\setminus A_j, B_i \setminus B_j)$, $(A_i \cap A_j, B_i \cap B_j)$ and
$(A_j \setminus A_i, B_j \setminus B_i)$.

To see it in terms of the theory explained in
Section~\ref{sec:surveycertalgo}, this results in the weak closure of
$\cF$. Let $\cF$, $\cI$ be the set system and interval
system (A set system where all the sets are intervals.),
respectively, involved in the assignment $\Pi$ and $\cF'$, $\cI'$ be
the same for the output of the algorithm $\Pi'$.  It can be observed
that $\cF' \cup \cF$ is $\overline{\cF}$ or the weak closure of
$\cF$. The analog is true for $\cI \cup \cI'$.  Moreover, this output
$\cF'$, $\cI'$ contains the strong elements of $\overline{\cF}$,
$\overline{\cI}$ respectively. It turns out that $\cI'$ is also an
interval system and that $\Pi'$ is an ICPIA \cite[Lem.~2]{nsnrs09}.

@@

What we saw so far in this section is checking for the feasibility of
a given interval assignment. ICPIA can also be used to find an
interval assignment when only the set system $\cF$ is given. Part of
the approach is similar to the overlap component idea in \cite{fg65,
  wlh02}. Once the overlap components are factored out, they can be
independently assigned a subinterval to which the sets in the
component need to be mapped. This is simple because the overlap
components are disjoint from each other. Now the subproblem is finding
\COP order for each overlap component. In each component, the set
whose overlaps with the rest of the sets in that component forms a
single inclusion chain (see Definition~\ref{def:poset}. In this
particular reference, the set of sets is the set of overlaps.)  is
chosen first and assigned the leftmost\footnote{This is by
  convention. Rightmost can also be chosen analogously to get a
  different \COP order.} interval. The next candidate set is chosen
from this set's overlapping sets. The interval assigned is calculated
by simple intersection cardinality means. Assume that the next set
overlaps on one of the ``ends'' of the current set and calculate the
interval by shifting left or right of the current interval and
adjusting its size to match the next set's cardinality. Once the next
set is assigned an interval, the current assigned sets are all checked
for ICPIA. This is a backtracking algorithm -- if at any point ICPIA
fails, another overlapping set is chosen. If all overlapping set
attempts fail ICPIA, a different ``first set'' is chosen. This is a
remarkably simple algorithm due to its obvious simplicity in
implementation.

@@

\tnote[dire]{ abrupt: Add a section closing para. does this method (above)
  lead to efficient algo? }

\section{Matrices without \COP}
\label{sec:surveycopopt}

\COP is a very beneficial property since it simplifies the structure
of the input (binary matrix or set system) leading to
efficient algorithms to otherwise hard problems.  While
Section~\ref{sec:surveycoptest} discusses matrices with \COP and how
to compute \COP orders, this section describes a couple of problems
when the input does not have \COP and identify their forbidden
substructures that prevents them from having \COP.
%describes the problems  that are
% of interest when a matrix does not have this property

\subsection{Finding forbidden submatrices}
It is known from Theorem~\ref{th:tuckercop}
that when a matrix does not have \COP, it must have one of Tucker's
forbidden submatrices. However, none of the \COP test algorithms discussed in
Section~\ref{sec:surveycoptest} outputs the forbidden submatrix when
the test is negative. This certainly gives a hint to how hard it is to find
these substructures in the input. \tnote[dire]{Fill neg results on
  finding forbidden submartrices -- boo75, some highlighted stuff from v85}

@@

\begin{corollary}[{\cite[Cor.~3.1]{d08phd}}]
  \label{cor:tuckertheorem}
  A matrix $M$ has \COP \iff its representing bipartite graph $G_M$
  (the graph whose half adjacency matrix is $M$) does not contain an
  asteroidal triple whose three vertices correspond to columns of M.
\end{corollary}

@@

\tnote[dire]{how does it give clues
  about the $R$ node? See dom.}

@@

\tnote[dire]{ -- sect 4.1 in cite:d08phd has many results surveyed. hardness
  results, approx. results. results are usually for a class of
  matrices $(a,b)$ where number {\un}s in columns and rows are
  restriced to $a$ and $b$ . -- problem of flipping at most $k$
  entries of $M$ to make it attain COP. this is NP complete
  cite:b75-phd%
\\(1) scite:lb62 showed that interval graphs are chordal and
  AT-free. %
\\couple of lines refering to tucker's submatrices. refer
  earlier section.

@@
